---
title: "Openstreetmap-Based Aerodrome Layouts"
author: "RQ"
date: "14/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(osmdata)
library(sf)
library(ggspatial)
```

This file plays with OSM and extracts data for plotting aerodrome layouts. 
An initial step towards wrapping the code into function is done further down.
The APT_BBOX.xlsx in the data-ad-charts sub-folder provides a list for the airports and 
associated bounding boxes. 

```{r}
# look-up bounding box, e.g. http://bboxfinder.com/
# -0.501251,51.454987,-0.410099,51.480538
bb <- matrix( c(-0.501251,51.454987,-0.410099,51.480538), ncol = 2, byrow = FALSE )
colnames(bb) <- c("min", "max")
rownames(bb) <- c("x", "y")
bb



q <- bb %>% opq() %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_minimal()
```

```{r}
# bboxfinder.com EGGK lng/lat
bb_lonlat <- c(-0.223915,51.141733,-0.155851,51.165738)
apt_icao  <- "EGKK"
apt_name  <- "London Gatwick"

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_void() +
  labs(title = "EGKK")

fn <- paste(apt_icao, ".png", sep = "")
ggsave(filename = fn, path = "data-ad-charts")

```

```{r}
# bboxfinder.com EGLL lng/lat
bb_lonlat <- c(-0.500123,51.454208,-0.408771,51.482241)
apt_icao  <- "EGLL"
apt_name  <- "London Heathrow"

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_void() +
  labs(title = apt_icao)

fn <- paste(apt_icao, ".png", sep = "")
ggsave(filename = fn, path = "data-ad-charts")

```



```{r}
# bboxfinder.com LEMD lng/lat
bb_lonlat <- c(-3.603793,40.449028,-3.519997,40.546982)
apt_icao  <- "LEMD"
apt_name  <- "Madrid Baraja Rudolfo Suarez"

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_void() +
  labs(title = apt_icao)

fn <- paste(apt_icao, ".png", sep = "")
ggsave(filename = fn, path = "data-ad-charts")
```






```{r}
osm_apt <- function(.bb_lonlat, .title = NULL, .add_north = TRUE){

  q <- opq(bbox = .bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

gg<-ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_void()

# if(!is.null(.title)){
#   gg <- gg + labs(title = .title)
# }
# add north arrow with ggspatial
 if(!is.null(.add_north)){
   gg <- gg +
     ggspatial::annotation_north_arrow(height = unit(0.7, "cm"), width = unit(0.7, "cm"))
 }
# return aerodrome layout map
  return(gg)  
}

save_osm_apt <- function(.gg, .apt_icao, .format = ".png", .dir="./data-ad-charts/"){
  #filename
  fn <- paste(.apt_icao, .format, sep = "")
  # save plot
  ggsave(plot = .gg, filename = fn, path = .dir)
}

db_apt_chart <- function(.apt_icao, .bb_lonlat, .add_north = TRUE, show = TRUE, ...){
  message(paste0("aerodrome layout plot for ", .apt_icao))
  apt_plot <- osm_apt(.bb_lonlat, .add_north)
  save_osm_apt(apt_plot, .apt_icao)
  if(show){print(apt_plot)}
}
```

```{r}
# bboxfinder.com EHAM lng/lat
bb_lonlat <- c(4.703521,52.279948,4.816517,52.368546)
apt_icao  <- "EHAM"
apt_name  <- "Amsterdam Schiphol"

eham <- osm_apt(bb_lonlat, .add_north = TRUE)
# rotates, but it is unclear to me how to identify the best rotated "projection", i.e. crs:
# eham + coord_sf(crs = 3338)
save_osm_apt(eham, apt_icao)
eham

# maximise plotting area == set plot.margin to min ==> check, if titles and axis!
eham + theme(plot.background = element_rect(fill = "darkseagreen"), plot.margin = unit(c(-0.5,-0.4,-0.6,-0.4), "cm"))
```

```{r}
# bboxfinder.com EDDF lng/lat
bb_lonlat <- c(8.486218,49.991121,8.604321,50.054538)
apt_icao  <- "EDDF"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com EDDM lng/lat
bb_lonlat <- c(11.734463,48.335210,11.837803,48.373311)
apt_icao  <- "EDDM"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com EDDK
bb_lonlat <- c(7.099262,50.844941,7.183033,50.890756)
apt_icao  <- "EDDK"

db_apt_chart(apt_icao, bb_lonlat)
```



```{r}
# bboxfinder.com LFPG lng/lat
bb_lonlat <- c(2.480164,48.979000,2.655258,49.034177)
apt_icao  <- "LFPG"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com LIRF lng/lat
bb_lonlat <- c(12.205605,41.771517,12.285942,41.854168)
apt_icao  <- "LIRF"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com LEMD lng/lat
bb_lonlat <- c(-3.612373,40.447313,-3.521393,40.546000)
apt_icao  <- "LEMD"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com LSZH lng/lat
bb_lonlat <- c(8.519966,47.437307,8.581593,47.488255)
apt_icao  <- "LSZH"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com EIDW lng/lat
bb_lonlat <- c(-6.309792,53.416387,-6.228939,53.444308)
apt_icao  <- "EIDW"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com EGGK lng/lat
bb_lonlat <- c(-0.223915,51.141733,-0.155851,51.165738)
apt_icao  <- "EGKK"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# EGLL
bb_lonlat <- c(-0.501251,51.454987,-0.410099,51.480538)
apt_icao  <- "EGLL"

db_apt_chart(apt_icao, bb_lonlat)
```


```{r}
# bboxfinder.com WSSS lng/lat
bb_lonlat <- c(103.963623,1.310692,104.037609,1.394268)
apt_icao  <- "WSSS"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com LGAV lng/lat
bb_lonlat <- c(23.911370,37.910234,23.989497,37.964916)
apt_icao  <- "LGAV"

db_apt_chart(apt_icao, bb_lonlat)
```


## Start to industrialise

```{r}
# bboxfinder.com EKCH lng/lat
bb_lonlat <- c(12.596169,55.585234,12.681484,55.633229)
apt_icao  <- "EKCH"

db_apt_chart(apt_icao, bb_lonlat)
```

```{r}
# bboxfinder.com ELLX lng/lat
bb_lonlat <- c(6.179638,49.613101,6.243238,49.640456)
apt_icao  <- "ELLX"

db_apt_chart(apt_icao, bb_lonlat)
```


```{r}
apdf_apts <- readxl::read_excel("./data-ad-charts/APT_BBOX.xlsx", sheet = "Monitored")

layout_done <- list.files(path = "./data-ad-charts/", pattern = "[A-Z]{4}\\.png") %>%
  strtrim(width = 4)

todo_apts      <- apdf_apts %>% filter(!AIRPORT %in% layout_done) %>%
  select(AIRPORT, BBOX)

bb_coerce <- function(.bb_lnglat_string){
  bb <- strsplit(.bb_lnglat_string, ",")
  bb <- unlist(bb)
  bb <- as.numeric(bb)
  return(bb)
}

# rm(apt_icao)

# loop
# db_apt_chart(apt_icao, bb_lonlat,.add_north = TRUE)
# lnglats %>% purrr::pmap(.f=~print(paste0("APT: ", ..1, "BBOX: ", ..2)) )
todo_apts %>% purrr::pmap(.f=~db_apt_chart(.apt_icao = ..1, .bb_lonlat = bb_coerce(..2)))
```

this worked: db_apt_chart("EDDH", c(9.955158,53.613081,10.011978,53.659594
+ ),.add_north = TRUE)
check how BBOX is provided might not be a vector

## Studying Extracting Information from the OSM Data Query

```{r}
# bboxfinder.com EBBR lng/lat
bb_lonlat <- c(4.438580,50.883263,4.529560,50.920819)
apt_icao  <- "EBBR"
apt_name  <- "Brussels Airport"

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() # %>%   unique_osmdata()

(  ebbr <- db_apt_chart(apt_icao, bb_lonlat)  )
```

querried OSM data are stored in object q.

```{r}
q
```

```{r}
ebbr_pts <- q$osm_points
ebbr_pts
```

The sf-collection osm_points comes with further slots: i.e. osm_id, aeroway

```{r}
table(ebbr_pts$aeroway, useNA = "ifany")
```

The point-collection comprises 70 (named) gates.

```{r}
pts <- ebbr_pts %>% select(aeroway, ref)
pts
```

```{r}
table(pts$ref, useNA = "ifany")
```

The ref-slot seems to comprise the gate identifier.

```{r}
ebbr_gates <- ebbr_pts %>% select(aeroway, ref, geometry) %>%
  filter(! is.na(ref))

ebbr_gates
```

```{r}
ebbr + geom_sf(data = ebbr_gates # %>% filter(aeroway == "runway"),
               , colour = "red"
               , label)
```

## Add North Arrow

```{r}
# bboxfinder.com EDDS lng/lat
bb_lonlat <- c(9.182562,48.679163,9.251656,48.698644)
apt_icao  <- "EDDS"

# bboxfinder.com LEMD lng/lat
bb_lonlat <- c(-3.612373,40.447313,-3.521393,40.546000)
apt_icao  <- "LEMD"

osm_apt(bb_lonlat) +
  ggspatial::annotation_north_arrow(height = unit(0.7, "cm"), width = unit(0.7, "cm"))
```

## Trouble Shooting 

### EYVI

```{r}
# bboxfinder.com EYVI lng/lat
bb_lonlat <- c(25.269670,54.615076,25.305119,54.655414)
apt_icao  <- "EYVI"

# bboxfinder.com EDDS lng/lat
#bb_lonlat <- c(9.182562,48.679163,9.251656,48.698644)

# bboxfinder.com LEMD lng/lat
#bb_lonlat <- c(-3.612373,40.447313,-3.521393,40.546000)
#apt_icao  <- "LEMD"

#xx <- osm_apt(bb_lonlat, .add_north = TRUE)
#db_apt_chart(apt_icao, bb_lonlat)

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

gg <- ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE, color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_multipolygons
          ,inherit.aes = FALSE, color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  geom_sf(data = q$osm_polygons %>% filter(aeroway == "runway"),
          inherit.aes = FALSE, fill = "black"
          #, color = "black", size = 2 #.4
          ,alpha = .8) +
  theme_void()
gg +
     ggspatial::annotation_north_arrow(height = unit(0.7, "cm"), width = unit(0.7, "cm"))

save_osm_apt(
   .gg = gg +
     ggspatial::annotation_north_arrow(height = unit(0.7, "cm"), width = unit(0.7, "cm"))
  ,.apt_icao = "EYVI")
```

### EDDR

```{r}
# EDDR bounding box
bb_lonlat <- c(7.092190,49.212860,7.126050,49.220849)
apt_icao  <- "EDDR"

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() %>%
  unique_osmdata()

eddr <- osm_apt(bb_lonlat, .add_north = TRUE) 
eddr <- eddr + 
  geom_sf(data = q$osm_polygons %>% filter(aeroway == "runway")
          ,fill = "black")
#db_apt_chart(apt_icao, bb_lonlat)



gg <- ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE, color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_multipolygons
          ,inherit.aes = FALSE, color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  geom_sf(data = q$osm_polygons %>% filter(aeroway == "runway"),
          inherit.aes = FALSE, fill = "black"
          #, color = "black", size = 2 #.4
          ,alpha = .8) +
  theme_void()
gg +
     ggspatial::annotation_north_arrow(height = unit(0.7, "cm"), width = unit(0.7, "cm"))

save_osm_apt(
   .gg = gg +
     ggspatial::annotation_north_arrow(height = unit(0.7, "cm"), width = unit(0.7, "cm"))
  ,.apt_icao = "EYVI")
```

